---
title: "Suicide Risk Predictors in Shandong, China — EDA & Logistic Regression"
author: "Carmen Roland"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
options(scipen = 999)
```

## Overview

This analysis explores predictors of the suicide-attempt outcome (`Died`: yes/no) in the Shandong suicide attempts dataset, and fits a binomial logistic regression model, with associated diagnostics.

**Variables of interest:** `Died` (outcome); `Urban`, `Age`, `Sex`,
`Education`, `Method` (predictors).

## Libraries

```{r libraries}
library(tidyverse)
library(Hmisc)
library(DEGreport)
library(lmtest)
library(alr4)
library(car)
library(RcmdrMisc)
library(pROC)
library(performance)
library(see)
library(DHARMa)
library(glmtoolbox)
library(MASS)
library(e1071)
library(caret)
```

## Data Import (Set directory)

```{r import}
data <- read.csv("SuicideChina - Working File - Copy.csv")

head(data)

# Summarize dataset
summary(data)

# Display rows and columns
dim(data)
```

```{r missing-values}
# Missing values per column
sapply(data, function(x) sum(is.na(x)))
```

```{r frequency-tables}
table(data$Died)
table(data$Urban)
table(data$Sex)
table(data$Education)
table(data$Method)
```

## Univariate Visualizations

```{r plot-died}
ggplot(data = data, aes(x = Died)) +
  geom_bar(fill = "red", color = "black") +
  ggtitle("Outcome of Suicide Attempt")
```

```{r plot-urban}
ggplot(data = data, aes(x = Urban)) +
  geom_bar(fill = "red", color = "black") +
  ggtitle("Location of Suicide Attempt")
```

```{r plot-age}
ggplot(data = data, aes(x = Age)) +
  geom_histogram(fill = "red", color = "black") +
  ggtitle("Age Distribution")
```

```{r plot-sex}
ggplot(data = data, aes(x = Sex)) +
  geom_bar(fill = "red", color = "black") +
  ggtitle("Sex Distribution")
```

```{r plot-education}
ggplot(data = data, aes(x = Education)) +
  geom_bar(fill = "red", color = "black") +
  ggtitle("Education")
```

```{r plot-method}
ggplot(data = data, aes(x = Method)) +
  geom_bar(fill = "red", color = "black") +
  ggtitle("Method")
```

## Bivariate Visualizations (Died vs. Predictors)

```{r bivariate-urban}
ggplot(data = data, aes(x = Died, fill = Urban)) +
  geom_bar(position = "dodge") +
  ggtitle("Number of Deaths/Attempts by Location of Attempt")
```

```{r bivariate-age}
ggplot(data = data, aes(x = Age, fill = Died)) +
  geom_bar(position = "dodge") +
  ggtitle("Age and Number of Deaths/Attempts")
```

```{r bivariate-sex}
ggplot(data = data, aes(x = Died, fill = Sex)) +
  geom_bar(position = "dodge") +
  ggtitle("Number of Deaths/Attempts by Sex")
```

```{r bivariate-education}
ggplot(data = data, aes(x = Died, fill = Education)) +
  geom_bar(position = "dodge") +
  ggtitle("Number of Deaths/Attempts by Education")
```

```{r bivariate-method}
ggplot(data = data, aes(x = Died, fill = Method)) +
  geom_bar(position = "dodge") +
  ggtitle("Number of Deaths/Attempts by Method")
```

## Data Preparation for Modeling

```{r prep-data}
df <- data.frame(
  died      = factor(data$Died),
  urban     = factor(data$Urban),
  sex       = factor(data$Sex),
  age       = data$Age,
  education = factor(data$Education),
  method    = factor(data$Method)
)

head(df)

# Set reference levels
df$urban_RL     <- relevel(df$urban, ref = "unknown")
df$education_RL <- relevel(df$education, ref = "unknown")
df$Method_RL    <- relevel(df$method, ref = "unspecified")
```

## Logistic Regression Model

```{r fit-model}
fit <- glm(
  died ~ urban_RL + age + education_RL + sex + Method_RL,
  family = binomial,
  data   = df
)

summary(fit)
Anova(fit)
```

## Confidence Intervals

Computed directly from the fitted model (not hardcoded), on both the log-odds and odds-ratio scales.

```{r confint}
ci_log_odds <- confint(fit, level = 0.95)
ci_log_odds

or_table <- cbind(
  OR = exp(coef(fit)),
  exp(ci_log_odds)
)
or_table
```

## Model Checking

```{r hltest}
# Hosmer-Lemeshow test — high p-value indicates adequate fit
hltest(fit)
```

```{r stepwise}
# Stepwise regression — sample is large enough; nothing dropped
stepwise(fit, direction = "backward/forward", criteria = c("BIC"))
stepwise(fit, direction = "backward/forward", criteria = c("AIC"))  # matches BIC
```

```{r performance}
model_performance(fit)
```

## Simulation-Based Residual Diagnostics (DHARMa) 

Note: Generalized linear model residuals are not expected to follow a normal distribution and require simulation-based diagnostics.

```{r dharma-setup}
plotConventionalResiduals(fit)
testDispersion(fit)

simulation_output <- simulateResiduals(fittedModel = fit, plot = FALSE)
residuals(simulation_output)
```

```{r dharma-plots}
plotQQunif(simulation_output)
plotResiduals(simulation_output)
plotResiduals(simulation_output, form = df$age)
```

```{r dharma-categorical-tests}
testCategorical(simulation_output, catPred = df$urban)
testCategorical(simulation_output, catPred = df$education)
testCategorical(simulation_output, catPred = df$Method_RL)
testCategorical(simulation_output, catPred = df$sex)
```

```{r dharma-global-tests}
testUniformity(simulation_output)
testOutliers(simulation_output)
testDispersion(simulation_output)
testQuantiles(simulation_output)
```

## ROC Curve

```{r roc}
fitted_probs <- predict(fit, type = "response")
roc_curve    <- roc(df$died ~ fitted_probs)
roc_curve$auc

plot(
  1 - roc_curve$specificities, roc_curve$sensitivities,
  type = "l", lwd = 2,
  xlab = "1 - Specificity", ylab = "Sensitivity"
)
abline(a = 0, b = 1)
```

## Interaction Model

Method is excluded from interactions (it causes overfitting); interactions were not significant regardless.

```{r interaction-model}
fit_interaction <- glm(
  died ~ urban_RL + age + education_RL + Method_RL +
    urban_RL * age + urban_RL * education_RL + age * education_RL,
  family = binomial,
  data   = df
)

summary(fit_interaction)

lrtest(fit_interaction, fit)  # not significant — simpler model preferred
```

## Classification / Confusion Matrix

```{r confusion-matrix}
prop_died  <- 1256 / 2571
prop_lived <- 1315 / 2571
prop_died
prop_lived

df$outcome <- ifelse(df$died == "yes", 1, 0)
head(df)

my_steps    <- step(fit, died ~ urban_RL + age + education_RL + sex + Method_RL,
                     data = df, family = "binomial")
fitted_vals <- fitted(my_steps)

confusionMatrix(
  data      = as.factor(as.numeric(fitted_vals > 0.5)),
  reference = as.factor(df$outcome)
)
```

## Key Findings

- The following predictors significantly predict suicide: sex (female), location of residence (rural), age, highest education level and method of attempt (except for other poison, and poison with type unspecified.) 
- All predictors independently predict suicide death and are not modified by a third predictor (an interaction.)
- The final logistic regression model achieved an AUC of 0.8936.

 